Moisture absorption and drying alter nonadditive litter flammability in mixed conifer forest

### 2019-12-20 ms/response_to_reviews.tex

\name{Dylan W. Schwilk}

\begin{document}
\begin{letter}{}

\opening{Dear Dr. David Gibson,}

Enclosed is our revised manuscript research article, ``Moisture absorption and drying alter nonadditive litter flammability in mixed conifer forest''.

### 2019-12-20 ms/response_to_reviews.tex

We thank the editors and reviewers for their substantial comments and
suggestions. We have itemized each comment below and followed each with our
reply and a description of changes we made in response to the comment.

{\bf Responses to the Senior Editor's comments:}

\begin{quoting}
  I support the AE's recommendation to reject but allow resubmission. However,
  the bar for a resubmission will be high. My feeling is that your study
  confirms the essential findings from your 2012 study published in Journal of
  Ecology with the addition that your methodology of drying down adds some
  realism. Indeed, the confirmatory nature of this study was pointed out by
  reviewer 2 in their point 6. In addition to the points raised by the AE and
  reviewers, I will be looking in any resubmission for more evidence that this
  study advances your earlier work (apart from your last paragraph, your
  discussion is largely an explanation of your results with little context from
  other studies), that the findings have real ecological relevance (i.e., does
  it matter to the mixed-conifer communities?) and how is the new dry down
  treatment related to dry down conditions in litter mixtures in the field. If
  you can adequately address all of these concerns then I'll be happy to
  consider a resubmission.
\end{quoting}

We agree that our Discussion was weak and we have rewritten it. We argue,
however, that this work is not merely confirmatory. Our revised Introduction
argues for the ecological significance of drying rates. The importance of fuel
moisture is well established. At large scales, fire in the western US is often
limited by fuel moisture (Parks et al 2014). At fine scales, variation in fuel
moisture influences heterogeneity in fire behavior (Hille and Stephens 2005)
and may contribute to fire history variation over short distances (Stephens
2001). Although the importance of drying rate is less well-established,
prescribed burns in Sierra Nevada mixed conifer forest demonstrate that fuel
moisture drives differences in fire behavior depending on the timing of fire
throughout the season (Kaufman and Martin 1989, Knapp and Keeley 2006, Hille
and Stephens 2005, Knapp et al 2006). These differences in fire behavior cause
spatial variation in tree mortality (Stephens et al 2002, Schwilk et al 2006).
Despite this, we know of only one other study specifically investigating
differential drying behavior (Kreye et al 2013b) and none that included drying
rates with flammability measurement.

@@

Thank you for the consideration and comments. Several reviewer comments
converged on the correct criticism that our Discussion did not engage
adequately with the literature. We have made substantial changes to the
Introduction and the Discussion adding more context and citing more of the
recent literature. We reorganized the Introduction and now do not rush the
reader so quickly through the ideas. We have expanded our justification for the
study as argued for by the Senior Editor and others.

\begin{quoting}
  The authors mention leaf size explicitly as an important parameter in
  non-additivity (with some references including their own 2012 paper) but they
  do not include it in the current study. Especially at low moisture, the small
  individual needles of Abies should cause low ventilation and a high
  likelihood of negative non-additivity on flammability, as was shown by Zhao
  et al 2016 Ecol. Evol. Figure 4).
\end{quoting}

We have added a new figure in response to this comment (SI Fig 1).

When considering how to limit the number of tests at the comparative scale
(species as data points), we decided to not include size, because its influence
on flammability is primarily through litter packing and was captured by
litterbed density. However, in response to the request to show particle sizes,
we have added a figure (SI Fig. 1) to the SI materials showing litter particle
size distributions for each of these species. These data were collected from
the stored litter. We did not add these data to the main figures as they are
not part of any statistical test or main conclusions, but we are happy to do so
if the editor suggests.

@@

\begin{quoting}
line 290 - you show an association, not necessarily an effect
\end{quoting}

We have edited as suggested (line 324)

\begin{quoting}
  line 307 and line 326 - a minor inconsistency is that you analyse present
  species groups, yet discuss the results in terms of individual species
\end{quoting}

We've tried to increase consistency throughout. Yes, we examined dry down
curves by species because we did not have prior hypothesis about groups and
because we had decided to find moisture dynamic groups empirically for deciding
on mixtures. For the flammability analyses, we had decided to examine patterns
by taxonomic group because we thought that would allow clearer presentation and
aid generalizability and understanding. We now refer to taxonomic groups when
appropriate and have standardized the subsection headings. We could do all
analyses on a species basis, but that would be re-running new models on those
data.

@@

\begin{quoting}

  My major concerns are:

  1. One motivation for the study is not valid. I do not completely agree with
  the authors that natural lightning ignitions are concurrent with
  precipitation and, therefore, fire behavior can depend on rate of drying
  following a wetting event. The efficiency of individual lighting strikes in
  igniting a fire is affected by the moisture properties of forest fuel at the
  moment of lighting (Krawchuk et al. 2006, Ecology). Dry lighting that occurs
  without any accompanied rainfall nearby is most likely to cause wildfire
  because the fuel is dry enough to be ignited.
\end{quoting}

Our sentence was poorly written and oversimplified. We thank the reviewer for
pointing this out and we have rewritten this section. However, our badly made
point was: lighting ignitions can occur during the same general period as
precipitation; therefore, drying rates are important. As the reviewer points
out, dry lightning strikes that find dry fuel are more likely to lead to fires
than are strikes that find damp fuels (Abatzoglou et al 2016). The fact that
the total number lightning strikes does not well predict burned area and that
observed ignitions occur most often from dry strikes argues for the importance
of fuel moisture. Although synoptic weather conditions that create dry and wet
storms differ, the potential for precipitation and lightning during the same
overall period of thunderstorm activity often exists (eg wet and dry convection
days can occur under overlapping conditions, Rorig and Ferguson 1999). We've
rewritten this section and added details justifying the importance of drying
rates in fine fuels as suggested by the Senior Editor. See lines 100--108 and
other changes to the Introduction.

@@

\begin{quoting}
  3. The authors concluded that positive non-additivity appeared to be the rule
  in litter-driven fire and this non-additivity became more positive with fuel
  drying. But if we look closely at figure 8, non-additivity was negative
  within 50 hours since wetting, non-additivity was both positive and negative
  at about 75 hours since wetting, and non-additivity was positive at about 100
  hours since wetting. The high variability of non-additive effects at about 75
  hours since wetting might have important ecological effects on fire ignition
  and behavior, and are important information cannot be ignored.
\end{quoting}

This is a good point and we have added some discussion of this pattern. We
tried to limit our statements to those supported by the statistical tests.
This, of course, obscures some details which are worth mentioning and we have
edited this section in response to the reviewer's criticism. We agree that the
high variability in nonadditivity across mixtures deserves highlighting and
that is the reason we presented the actual predictions and departures from
prediction across time. Please see our revised Discussion and lines 462--474.

\begin{quoting}
  4. The writing of this manuscript needs to be improved for clarity and
  accuracy. There are many typos in the text, and the using to scientific terms
  lack consistency through the text.
\end{quoting}

We have made extensive edits in response to both reviewer's suggestions.
Furthermore, we have standardized our usage of some terms. We use the term
``maximum water absorption'' to indicate water content following saturation. Of
course, this is slightly inaccurate and includes both absorption into particles
and adsorption to surfaces. Our earlier draft used the term ``maximum water
retention''. But we think this earlier wording was more confusing (and we were
inconsistent in usage as the reviewer points out).

@@

\begin{quoting}
Line 50: add reference to the first sentence.
\end{quoting}

We have added one.

\begin{quoting}
  Lines 50-61: in this paragraph, after the first sentence I think it is more
  logic to first explain the mechanism of moisture’s negative effects on fire,
  then explain the reason of the driving importance of dead fuel moisture
  content to surface fire ignition and behavior because fine dead fuels respond
  more readily to environmental changes than living fuels.
\end{quoting}

We have rearranged the logic of the Introduction as suggested.

@@

\begin{quoting}
Line 85: revise ``driven by by'' and ``emergent litter litter''.
\end{quoting}

Fixed.

\begin{quoting}
Line 88: “similarly”?
\end{quoting}

We were not sure what the reviewer's criticism was here but have rewritten the
Introduction.

@@

\begin{quoting}
  Line 185: I do not think this sentence is necessary because whether or not
  petioles should be included in SLA measurements depends on the research
  question at hand.
\end{quoting}

Agreed and done.

\begin{quoting}
  Line 201: the fuel depth was controlled or not?
\end{quoting}

In flammability trials mass was fixed, width and length of the burning tray was
fixed (15\,cm by 1.5\,m), and depth was allowed to vary. We edited this section
for clarity. This decision was based on discussion with materials engineers who
study fire that recommend a burning tray length of 10 times width to best
measure steady flame spread. See lines 236--250.

@@

\begin{quoting}
  Line 216: revise “mass loss” to “fuel consumption”.
\end{quoting}

Done.

\begin{quoting}
  Line 224: if not all flammability data was used in the analysis, then why
  measure them?
\end{quoting}

Flammability is a multivariate trait. To avoid collinearity and to simplify
models, two general approaches are possible: either reduce and transform the
response variables and create synthetic variables through a method such as PCA,
or pick a few variables representative of the axes of variation. When the
experiment was designed, we had intended to use the first method. The loss of
several days of thermocouple data made that an unattractive and unnecessary
option, so we opted for the simpler method of picking two traits already known
to correlate with different axes of litter flammability.

We could eliminate any mention of the other measurements if the reviewer and
editors wish. In this draft have erred towards full disclosure.

### 2019-12-20 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Oops, typo in title fixed.
Fix typos

\title{Moisture absorption and drying alters nonadditive litter flammability
  in mixed conifer forest}

@@

      \item Moisture content is a strong determinant of forest fuel
    flammability, however, little is known about the mechanisms by which
    species-specific litter traits influence litter moisture absorption and
    retention. Characterizing how litter fuels retain moisture through time and
    how fuels from multiple species interact are key steps to adding species or
    taxon specific information to fire behavior prediction.

@@

      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. We examined two axes of
    flammability variation and found that these differing moisture dynamics
    result in time since wetting having strongly differing effects on
    flammability across the major litter types typified by different tree taxa.

>>>>

\title{Moisture absorption and drying alter nonadditive litter flammability
  in mixed conifer forest}

@@

      \item Moisture content is a strong determinant of forest fuel
    flammability, however, little is known about the mechanisms by which
    species-specific litter traits influence litter moisture absorption and
    retention. Characterizing how litter fuels retain moisture through time and
    how fuels from multiple species interact are key steps to adding taxon
    specific information to fire behavior prediction.

@@

      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. We examined two axes of
    flammability variation and found that these differing moisture dynamics
    resulted in time since wetting having strongly differing effects on
    flammability across the major litter types typified by different tree taxa.
